---
title: "Data Science Capstone - Principal Componenet Analysis"
author: "Nick Belgau, Oscar"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction (abstract?)

Significance + scope of the paper
PCA is best used when there is reason to believe that some underlying, latent factors explain the patterns in the data.
After performing PCA, validate the practical utility of the reduced dimensions by applying them in downstream tasks such as clustering, classification, or regression. Check if the reduced data maintains or improves the performance of these tasks.




## Methods

### Linear Algebra Foundations

Mathematical underpinnings
1. Standardize the data
2. Compare covariance matrix
3. Perform eigen decomposition or SVD. Each axis is orthogonal (no correlation). This helps eliminate multicollinearity. May not hold true if original variables are statistically uncorrelated.
4. Sort eigenvalues DESC, each is a principal component
5. Create projection matrix and project onto new feature space

Computations techniques. methods for computing PCA like SVD

Cite this paper [@bro2014principal]. 


### PCA Assumptions

1. Assumption: Linearity
Assume data is represented as linear combinations of each other.
check for non-linear relationships by using visual methods and statistical methods.
visually: correlation matrix where each 

corr plot:
```{r}SCATTER PLOT MATRICES: TO CHECK LINEARITY ASSUMPTION
par(
  col.axis="#8B814C", col.lab="#8B814C", col.main="#8B814C", col.sub="#8B814C", pch=20, col="#8B814C", bg="transparent"
)
DescTools::PlotPairs(sapply(conspiracy[,-9], function(x) jitter(x, 5)), 
                     g=conspiracy$y,
                     col=alpha("#8B814C", 0.1), 
                     col.smooth="#8B814C")
```
If the plot shows nonlinear relationships, need to examine

Look at distribution plots with overlayed mean=0 and std dev=1. and use it to example scaling.

Can we use statistical tests for linearity like Breusch-Pagan test or the Randombow test.
Or in practice do we use Pearson correlation coefficient?
Can apply transformation to convert from a non-linear relationships, and reassess statistical test.

Or use a different method like t-SNE or UMAp if PCA does not seem suitable


2. Assumption: PC with the highest variance are the most informative.
This is based on the idea that high variance directions capture more significant behaviors and structures in the data.
Outliers or noise in the data can lead to misleading conclusions because they may not necessarily be the most 'informative' part of the data.


Show what happens when you dont scale the data.

### Best Practices

1. Preprocessing - Data scaling
Assess the range, mean , and variance.
Variables on larger scales can disproportionately influence the outcome. Standardie at mean = 0 variance = 1 Show what happens when you dont scale.

2. Preprocessing - examine outliers

## Applications: Analysis and Results

### Application 1: General Dimensionality Reduction

Screeplot
Biplot

In one of these applications, discuss how PCA aids in building more robust statistical models by reducing multicollinearity and enhancing model interpretability.
Compare VIF (variance inflation factor) before and after PCA.


### Application 2: Feature Extraction
continuation of application 1 but go into ML part

### Application 3: Image Compression

New example.
Probably dont need to standardize the data. 
No scree plot or biplot because pixels.
Results to focus on memory reduction and maintaining equivalent level of performance.
This will be similar to appl 1 + 2 but without going into as much detail for techniques already demonstrated.

## Conclusion

Summary, reflections, future discissions

## References





```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
```

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$