---
title: "Principal Component Analysis"
author: "Shivani"
format: html
course: STA 6257 - Advanced Statistical Modeling
editor: visual
---

**Dataset**

Each record is for 100 grams.

The nutrient columns end with the units, so:

Nutrient_g is in grams Nutrient_mg is in milligrams Nutrient_mcg is in micrograms Nutrient_USRDA is in percentage of US Recommended Daily Allows (e.g. 0.50 is 50%)

#Install Libraries

```{r}
#install.packages("devtools")
#library(devtools)
#install_version("MASS", version = "7.3-60")  # Replace with a compatible version

#install.packages("MASS")
```

**Check for highly correlated features**

Removing _USRD records as the data is redundant and also selecting highly corrrelated data.

```{r}
library(MASS)
library(factoextra)
library(ggplot2)
library(readr)

library(dplyr)
library(caret)
library(tibble)

url <- "https://query.data.world/s/ll77ildgnhhove7mlker3g2jw7z5qr?dws=00000"
data <- read.csv(url, header=TRUE, stringsAsFactors=FALSE)
str(data)
#summary(data)


data_omit_usrda <- data %>% dplyr::select(-contains('_USRDA'))

str(data_omit_usrda)

data_numeric <- data_omit_usrda %>% dplyr::select(where(is.numeric))

cor(data_numeric)

head(data_numeric, 15)



corrs <- cor(data_numeric)
corrsdf <- as.data.frame(as.table(corrs))
corrsdf <- corrsdf %>% filter(Freq > 0.9 & Var1 != Var2)
head(corrsdf, 15)


df <- data_numeric %>% column_to_rownames('ID')
df_desc <- df[, 1:6]
df_new <- df[, -c(1:6)]


```

Checking features dustribution 

```{r}
#install.packages("tidyr")
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

df_new %>% gather() %>% ggplot(aes(value)) + 
  geom_histogram(bins=50) + 
  facet_wrap(~key, scales='free') + 
  theme_minimal()
```

**Checking data variance**

```{r}
df <- scale(df_new)
cat("mean:", round(mean(df), 2), "\n")
cat("standard dev:", round(sd(df), 2), "\n")

```
```{r}
pca <- prcomp(df, scale = TRUE)
qplot(1:length(pca$sdev^2), pca$sdev^2 / sum(pca$sdev^2), geom="line")

explained_variance <- pca$sdev^2 / sum(pca$sdev^2)
cat(explained_variance, "\n")
cat(sum(explained_variance[1:5]), "\n")


```

```{r}
pca_df <- as.data.frame(pca$x[, 1:5])
pca_df <- rownames_to_column(pca_df, var = "ID")
#pca_df <- left_join(pca_df, df_desc, by="ID")
#pca_df <- pca_df %>% select(-CommonName, -MfgName, -ScientificName)
names(pca_df)[2:6] <- c('c1', 'c2', 'c3', 'c4', 'c5')
names(pca_df)
round(cor(pca_df[, 2:6]), 5)

```

```{r}
vects <- pca$rotation[, 1:5]

one <- sort(vects[, 1], decreasing = TRUE)
two <- sort(vects[, 2], decreasing = TRUE)
three <- sort(vects[, 3], decreasing = TRUE)
four <- sort(vects[, 4], decreasing = TRUE)
five <- sort(vects[, 5], decreasing = TRUE)

```






