---
title: "Shivani_PCA_Presentation"
author: "Shivani"
format:
  revealjs:
    toc: true
    smaller: true
    progress: true
    slide-number: true
    width: 1280
    height: 960
editor: visual
---

### Dataset Description

-   The USDA National Nutrient Database for Standard Reference (SR) is the major source of food composition data in the United States.

-   It provides the foundation for most food composition databases in the public and private sectors.

-   As information is updated, new versions of the database are released.

-   Each record is for 100 grams.

-   The nutrient columns end with the units, so: Nutrient_g is in grams Nutrient_mg is in milligrams Nutrient_mcg is in micrograms Nutrient_USRDA is in percentage of US Recommended Daily Allows (e.g. 0.50 is 50%)

---

### Load Data and transform if needed

- Load data
- Transform dataset removing NULLs and any non numerical features

```{r}
library(MASS)
library(factoextra)
library(ggplot2)
library(readr)

library(dplyr)
library(caret)
library(tibble)

url <- "https://query.data.world/s/ll77ildgnhhove7mlker3g2jw7z5qr?dws=00000"
data <- read.csv(url, header=TRUE, stringsAsFactors=FALSE)
#str(data)
#summary(data)
data_omit_usrda <- data %>% dplyr::select(-contains('_USRDA'))

#str(data_omit_usrda)

data_numeric <- data_omit_usrda %>% dplyr::select(where(is.numeric))

df <- data_numeric %>% column_to_rownames('ID')
df_desc <- df[, 1:6]
data_numeric <- df[, -c(1:6)]


print(data_numeric, 15)

```
---

### Scale Data

```{r}
data_scaled <- scale(data_numeric)
cat("mean:", round(mean(data_scaled), 2), "\n")
cat("standard dev:", round(sd(data_scaled), 2), "\n")

```



---

### Corelation check

- Check if features in the data are correlated

- High correlated variables are used for effective PCA  

```{r}
cor(data_scaled)
mean(cor(data_scaled))


```

---

### PCA

- Running PCA generated multiple Pricipal components

- First few components will be important.


```{r}
pca <- prcomp(data_scaled, scale = TRUE)
qplot(1:length(pca$sdev^2), pca$sdev^2 / sum(pca$sdev^2), geom="line")

explained_variance <- pca$sdev^2 / sum(pca$sdev^2)
cat(explained_variance, "\n")
cat(sum(explained_variance[1:7]), "\n")


# Print summary of PCA
#print(pca)


```
---

### Scree Plot

- The scree plot shows the proportion of variance explained by each principal component.
- This plot helps in deciding how many components to retain for further analysis.
- First seven components explain most of the variance, (> 70%) 


```{r}
# Visualize the PCA results
fviz_eig(pca, addlabels = TRUE, ylim = c(0, 50))

```

---

### Biplot

```{r}
fviz_pca_biplot(pca, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

--- 

### Visualization

```{r}
pca_results <- as.data.frame(pca$x)
pca_results$FoodGroup <- data$FoodGroup
plot_top_foodgroups <- function(pca_results, component, n = 10) {
  top_foodgroups <- pca_results %>%
    arrange(desc(!!sym(component))) %>%
    slice(1:n) %>%
    count(FoodGroup, sort = TRUE)
  
  ggplot(top_foodgroups, aes(x = reorder(FoodGroup, n), y = n)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(title = paste("Top Food Groups for", component),
         x = "Food Group", y = "Count")
}

# Plotting the top food groups for the first 5 components
for (i in 1:7) {
  component <- paste0("PC", i)
  print(plot_top_foodgroups(pca_results, component))
}

```

---

### Creating dataset with pricipal components

- Creating dataset with 7 PCs 

```{r}
pca_df <- as.data.frame(pca$x[, 1:7])
pca_df <- rownames_to_column(pca_df, var = "ID")
#summary(pca_df)
cor(pca_df[, 2:8])
mean(cor(pca_df[, 2:8]))

```

---

### Summary


Top Contributing Food Groups:

- The bar plots for each of the first five principal components indicate which food groups are most influential in each component. This helps in understanding the underlying structure of the data and identifying patterns or groups of foods that are similar in their nutritional profiles.

Dimensionality Reduction:

- By reducing the dataset to a few principal components, this helps in simplifying the dataset while retaining most of the important information.
